---
title: "Untitled"
author: "Jacob Haarala"
date: "12/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown



```{r}
game_goals <- read_csv("game_goals.csv")
season_goals <- read_csv("season_goals.csv")
top_250_career <- read_csv("top_250.csv")
```

## Including Plots


```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.

```{r}
season_goals %>%
  filter(goals > 60) %>%
  ggplot(season_goals, mapping = aes(player, goals, fill = player)) +
  geom_boxplot() +
  theme(axis.text.x = element_text(angle = -90, hjust = 0)) +
  theme(legend.position = "None") +
  scale_fill_manual(values = c("Alex Ovechkin" = "red4", "Alexander Mogilny" = "navy", "Bernie Nicholls" = "gray4", "Bobby Hull" = "red4", "Brett Hull" = "green4", "Jari Kurri" = "orangered", "Jaromir Jagr" = "yellow3", "Lanny McDonald" = "red3", "Luc Robitaille" = "darkorchid4", "Mario Lemieux" = "yellow2", "Mike Bossy" = "chocolate1", "Phil Esposito" = "gold", "Reggie Leach" = "darkorange",  "Steve Yzerman" = "red3", "Teemu Selanne" = "navyblue", "Wayne Gretzky" = "mediumblue"))
```


```{r}
season_goals %>%
  filter(player == "Alex Ovechkin") %>%
  tail(10) %>%
  ggplot(season_goals, mapping = aes(season, goals)) +
  geom_point(shape=21, color = "red", fill = "midnightblue") +
  geom_line(color="red", group = 1) +
  theme(axis.text.x = element_text(angle = -45, hjust = 0)) +
  ggtitle("Alex Ovechkin") +
  theme(plot.title = element_text(hjust = 0.5))
```

```{r}

top_250_career %>%
  filter(total_goals > 675) %>%
  ggplot(top_250_career, mapping = aes(raw_rank, total_goals)) +
  geom_point(col = "darkred") +
  gghighlight(raw_rank > 0 & raw_rank < 2 ,
              label_key = player,
              unhighlighted_colour = alpha("steelblue", 0.4)) +
  geom_point(col = "darkred", size = 2.5) +
  scale_x_continuous(breaks = round(seq(min(0), max(14), by = 1),1)) +
  scale_y_continuous(breaks = round(seq(min(0), max(900), by = 25),1)) 

top_250_career %>%
  filter(total_goals > 675) %>%
  ggplot(top_250_career, mapping = aes(raw_rank, total_goals)) +
  geom_point(col = "darkred") +
  gghighlight(raw_rank > 7 & raw_rank < 9,
              label_key = player,
              unhighlighted_colour = alpha("steelblue", 0.4)) +
  geom_point(col = "darkred", size = 2.5) +
  scale_x_continuous(breaks = round(seq(min(0), max(14), by = 1),1)) +
  scale_y_continuous(breaks = round(seq(min(0), max(900), by = 25),1)) 

```


```{r}
glimpse(top_250_career)
```



 scale_color_manual(values = c("Alex Ovechkin" = "red4", "Brendan Shanahan" = "red3", "Brett Hull" = "green4", "Gordie Howe" = "red3", "Jaromir Jagr" = "yellow3", "Luc Robitaille" = "red3", "Marcel Dionne" = "purple", "Mario Lemieux" = "yellow2", "Mark Messier" = "dodgerblue4", "Mike Gartner" = "mediumblue", "Phil Esposito" = "gold", "Steve Yzerman" = "red3", "Teemu Selanne" = "navyblue", "Wayne Gretzky" = "chocolate1"))
 

